---
title: "Reproducible Research: Peer Assessment 1"
author: "Benoit Séguin"
date: '2024-09-22'
output:
  html_document: default
  word_document: default
---

<br>
This document presents the steps necessary to analyze the activity data which shows the number of steps taken in 5 minute intervals
<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

To answer the question, we will first need to load the required libraries to execute the code

```{r Load Libraries}
library(dplyr)
library(tidyr)
library(lubridate)
```

<br>

Then we will load the activity.csv data using read.csv

```{r Load Data}
activity <- read.csv("activity.csv")
```

<br>

Finally, we will prepare the data to ensure we can analyze it properly
```{r Prepare Data}
activity$date = as.Date(activity$date,"%Y-%m-%d")
```

## What is mean total number of steps taken per day?

<br>


#### 1. Calculate the total number of steps taken per day, ignoring missing values as indicated.

<br>

```{r Dailysteps}
DailySteps <- activity %>% group_by(date) %>% summarize(Steps = sum(steps,na.rm = TRUE))

```

<br>

#### 2. Plot histogram

<br>

```{r DailyStepsHist}
StepFrequency <- DailySteps$Steps
hist(StepFrequency, breaks=25, xlim=c(0,25000), xlab = "Number of steps (1000 steps interval)")
```

<br>

#### 3. Calculate and report the mean and median of total number of steps

<br>

```{r StepsMeanMedian}
MeanDailySteps <- mean(DailySteps$Steps, na.rm = TRUE)
MedianDailySteps <- median(DailySteps$Steps, na.rm = TRUE)
```

<br>


The daily steps mean is:

```{r Answer Q1 Mean}
print(MeanDailySteps)
```
<br>

The daily steps median is:

```{r Answer Q1 Median}
print(MedianDailySteps)
```
<br>

<br>

## What is the average daily activity pattern?

<br>

#### 1. Timeseries plot of the average number of steps per interval

<br>

Sum the total number of steps per interval, then average by interval and plot, ignoring missing values as indicated.

```{r Timeseries}
IntervalSteps <- 
  activity %>% group_by(interval) %>% 
  summarize(mean = mean(steps, na.rm = TRUE))

plot(
  IntervalSteps$interval,IntervalSteps$mean,
  xlim=c(0,2400),
  main = "Average number of steps per daily interval",
  xlab = "Interval",
  ylab = "Average number of steps",
  type="l")
```

<br>

#### 2. Identify which 5 minute interval, on average, contains the maximum number of steps

<br>

Use the max function to identify all 5 minute intervals with the max average value. 

```{r MaxInterval Calc}
MaxInterval <- 
  IntervalSteps[IntervalSteps$mean == max(IntervalSteps$mean), ]$interval
```

<br>

The interval with the max average is:

```{r Print MaxInterval}
print(MaxInterval)
```
<br>

## Imputing missing values

<br>

#### 1. Calculate and report the total number of missing values

<br>

Use the function is.na to identify missing values


```{r Missing values}
sum(is.na(activity$steps))
sum(is.na(activity$date))
sum(is.na(activity$interval))
```
<br>

There are 2304 records with missing steps

<br>

#### 2. Impute interval mean for missing steps

<br>

Replicate Steps column and Impute interval mean to NA values
```{r replicate Impute interval mean}
activity$ImputedSteps <- activity$steps
activity <- activity %>% 
  group_by(interval) %>%
  mutate(
    ImputedSteps = ifelse(is.na(ImputedSteps), mean(ImputedSteps, na.rm = TRUE),ImputedSteps))
```

<br>

Confirm no NA
```{r Confirm no NA}
sum(is.na(activity$ImputedSteps))
```
#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

<br>

Create new dataset
```{r activityNew}
activityNew <- activity[,c("ImputedSteps","date","interval")]
names(activityNew)[1] <- "steps"
head(activityNew)
```

<br>

#### 3. Recreate Q1 with activityNew. Are there differences in Mean and Median total number of steps taken per day

<br>

Repurpose code from Q1 with activityNew to create new histogram

```{r DailyStepsHistNew}
DailyStepsNew <- activityNew %>% group_by(date) %>% summarize(Steps = sum(steps,na.rm = TRUE))
StepFrequencyNew <- DailyStepsNew$Steps
hist(StepFrequencyNew, breaks=25, xlim=c(0,25000), xlab = "Number of steps (1000 steps interval)")
```

<br>

Calculate and report mean and median total number of steps



The new daily steps mean and median are:

```{r New daily mean and median}
MeanDailyStepsNew <- mean(DailyStepsNew$Steps, na.rm = TRUE)
MedianDailyStepsNew <- median(DailyStepsNew$Steps, na.rm = TRUE)
print(MeanDailyStepsNew)
print(MedianDailyStepsNew)
```
<br>

The impact of imputing missing data on the estimates of total daily has increased both the mean and median

<br>

## Are there differences in activity patterns between weekdays and weekends?

<br>

#### 1. Add weekday to the activityNew dataset

<br>

```{r Add weekdays}
activityNew$Day <- weekdays(activityNew$date)
```

<br>

Create factor variable to identify if weekday or weekend
```{r Add factor variable for day type}
activityNew$Weekday <- factor(ifelse(wday(activityNew$date) %in% c(1,7),"Weekend","Weekday"))
```

<br>

#### 2. Timeseries panel plot

<br>

Time series panel plot for the 5 minute intervals and average number of steps taken, averaged accross all weekday or weekend days

```{r Calculate mean steps based on Interval and Weekday grouping then plot data}
WeekdaySteps <- 
  activityNew %>% group_by(Weekday,interval) %>% 
  summarize(mean = mean(steps, na.rm = TRUE))



par(mfrow = c(2, 1), mar=c(2,4,2,0.2))
plot(
  WeekdaySteps$interval[WeekdaySteps$Weekday == "Weekday"],
  WeekdaySteps$mean[WeekdaySteps$Weekday == "Weekday"],
  xlim=c(0,2400),
  ylim=c(0,250),
  main = "Weekday average number of steps per daily interval",
  xlab = "Interval",
  ylab = "Average number of steps",
  type="l")
plot(
  WeekdaySteps$interval[WeekdaySteps$Weekday == "Weekend"],
  WeekdaySteps$mean[WeekdaySteps$Weekday == "Weekend"],
  xlim=c(0,2400),
  ylim=c(0,250),
  main = "Weekend average number of steps per daily interval",
  xlab = "Interval",
  ylab = "Average number of steps",
  type="l")

```